Case for Support

### 2021-11-17 application/Je-S_sections.tex
merged

The main beneficiaries of this proposal will be experimental brain and animal behaviour labs worldwide who currently lack the highly technical expertise required to apply Machine Intelligence to their experiments. In fields like systems neuroscience, where progress is made essentially by exploring the vast unknown space of animal behaviour and wide-scale brain physiology, the need for redesign of experimental protocols and rapid analysis of large and complex datasets is constant and critical. While increasingly automated Machine Learning platforms are becoming available across many data science fields, they require deep expertise or training to construct and train new algorithms, or they are easy to deploy and use but are limited to very specific domains (e.g. face recognition, pose estimation). If successful, this proposal will entirely transform the landscape of Machine Intelligence applied to experimental lab research in biology, by empowering non-technical researchers to recombine and repurpose Machine Intelligence tools using a flexible visual programming environment and providing seamless integration with all the measurement, instrumentation and control packages already available in Bonsai.

Machine learning and computational neuroscience experts will also benefit from a standard platform to compare and benchmark different algorithms and stress test their tools against community datasets and experiments. The addition of new programming languages to develop and interface with Bonsai will also greatly expand the potential for application of their methods into new domains of research.

The scientific community will collectively benefit from gains in efficiency and reproducibility of deploying Machine Intelligence pipelines and allowing non-experts to compose existing algorithms in a flexible way to design entirely novel Machine Intelligence applications tailor-made for specific research questions. The proposed tools will generate cost savings to individual laboratories as well as agencies who fund them by drastically reducing duplicated software development efforts across research groups.

In addition to the research community, the standard analysis pipelines developed in this project may be applied to other domains of society where tracking, monitoring and understanding animal behaviour would yield benefits and improved outcomes. For example, Bonsai is being increasingly used in research institutions worldwide to monitor animal welfare by in-house technicians who are concerned about a broader range of environmental variables which might impact animal husbandry but which are beyond the ones which can be monitored in standard vivarium solutions. As a free and open-source software package, Bonsai would empower such users to prototype and design new animal monitoring systems on-site, without the need to engage in expensive software development to answer simple questions, and would lead to instantly improving the welfare of hundreds of thousands of research animals in biological research facilities.

>>>>

The main beneficiaries of the proposed resource will be experimental brain and animal behaviour labs worldwide who currently lack the highly technical expertise required to apply Machine Intelligence to their experiments. In fields like systems neuroscience, where progress is made by exploring the unknowns of animal behaviour and wide-scale brain physiology, the need for redesign of experimental protocols and rapid analysis of large and complex datasets is constant and critical. While increasingly-automated Machine Learning platforms are becoming available across many data science fields, they require deep expertise or training to construct and train new algorithms; or else they are easy to use but limited to very specific domains (e.g. face recognition, pose estimation). If successful, this proposal will empower an entirely new community of non-technical researchers to recombine and repurpose Machine Intelligence tools using a flexible visual programming environment and providing seamless integration with all the measurement, instrumentation and control packages already available in Bonsai.

Machine Learning and computational neuroscience experts will also benefit from a standard platform to compare and benchmark different algorithms and stress-test their tools against community datasets and experiments. The addition of new programming languages to develop and interface with Bonsai will also greatly expand the potential for application of their methods into new domains of research.

The scientific community will collectively benefit from gains in efficiency and reproducibility of deploying Machine Intelligence pipelines, and from allowing non-experts to compose existing algorithms in a flexible way to design entirely novel Machine Intelligence applications tailor-made for specific research questions. The proposed tools will generate cost savings to individual laboratories as well as funding agencies, by drastically reducing duplicated software development efforts across research groups.

In addition to the research community, the standard analysis pipelines developed in this project may be applied to other domains of society where tracking, monitoring and understanding animal behaviour would yield benefits and improved outcomes. For example, Bonsai is  increasingly used in research institutions worldwide to monitor animal welfare by in-house technicians who are concerned about a broader range of environmental variables which might impact animal husbandry but which cannot be monitored in standard vivarium solutions. As a free and open-source software package, Bonsai would empower such users to prototype and design new animal monitoring systems on-site, without the need to engage in expensive software development to answer simple questions. It could thus contribute to improving the welfare of hundreds of thousands of animals in research facilities, zoos or farms. More broadly, Bonsai may also be applied to quantify dynamics of objects in wider settings from drone or satellite imagery, to track and quantify the dynamics of crop growth, deforestation or other biological ecosystems.

### 2021-11-17 application/bonsai.tex
merged

Bonsai also supports the growing wave of foundational
open hardware initiatives, such as OpenEphys \citep{siegleEtAl17} and UCLA
Miniscope~\citep{caiEtAl16}, allowing these tools to be quickly combined and integrated into new experiments~\citep{buccinoEtAl18}.
%
Bonsai has been adopted in large neuroscience undertakings like the
International Brain
Laboratory\footnote{internationalbrainlab.com/}
and the Allen Institute for Brain
Science\footnote{https://alleninstitute.org/what-we-do/brain-science/}.
%
Key to the wide adoption is the reproducibility that Bonsai ensures.  Data acquisition and experimental control protocols can be 
replicated in any laboratory just by sharing a Bonsai configuration file. 
%
% LEAVE OUT FOR THIS GRANT:
% More recently, Bonsai has started to expand outside the domain of neuroscience
% into biomedical research and biotechnology tool development, and even outside
% academia into public outreach and education programs.

>>>>

Bonsai also supports the growing wave of foundational
open hardware initiatives, such as OpenEphys \citep{siegleEtAl17} and UCLA
Miniscope~\citep{caiEtAl16}, allowing these tools to be quickly combined and integrated into new experiments~\citep{buccinoEtAl18}.
%
Bonsai has been adopted in large neuroscience undertakings like the
International Brain
Laboratory \footnote{internationalbrainlab.com/}
and the Allen Institute for Brain
Science\footnote{https://alleninstitute.org/what-we-do/brain-science/}.
%
Key to the wide adoption is the reproducibility that Bonsai ensures.  Data acquisition and experimental control protocols can be 
replicated in any laboratory just by sharing a Bonsai configuration file. 
%
% LEAVE OUT FOR THIS GRANT:
% More recently, Bonsai has started to expand outside the domain of neuroscience
% into biomedical research and biotechnology tool development, and even outside
% academia into public outreach and education programs.

### 2021-11-17 application/communityDemand.tex
merged

% Bonsai has directly enabled high-impact research in neuroscience~\citep[e.g.,][]{soaresEtAl16,grundemannEtAl17,netoEtAl16,hagiharEtAla21,anguillonRodriguezEtAl21} and contributed to accelerating technical progress in the field by functioning as an integrator of open-source, community developed technologies \citep{itskovEtAl14,lopesEtAl21,kaneEtAl20}.
The need for a strong and
flexible ML package that can be easily integrated with
real-time experimental control workflows has been raised multiple times by the
Bonsai user community, both as specific problems around 
tracking and interpretation of animal behaviour, and as 
explicit proposals to create a ML
package\footnote{https://groups.google.com/g/bonsai-users/c/BZ3zOOdv\_1c/m/x6OP75frBQAJ}. Even
relatively modest contributions such as integrating support for specific models, such as DeepLabCut, or the more specific zebrafish feature-tracking package BonZeb, have attracted positive
feedback from the community, \citep[e.g.,][]{kaneEtAl20,guilbeaultEtAl21}.

>>>>

% Bonsai has directly enabled high-impact research in neuroscience~\citep[e.g.,][]{soaresEtAl16,grundemannEtAl17,netoEtAl16,hagiharEtAla21,anguillonRodriguezEtAl21} and contributed to accelerating technical progress in the field by functioning as an integrator of open-source, community developed technologies \citep{itskovEtAl14,lopesEtAl21,kaneEtAl20}.
The need for a strong and
flexible ML package that can be easily integrated with
real-time experimental control workflows has been raised multiple times by the
Bonsai user community, both as specific problems around 
tracking and interpretation of animal behaviour, and as 
explicit proposals to create a ML
package\footnote{groups.google.com/g/bonsai-users/c/BZ3zOOdv\_1c/m/x6OP75frBQAJ}. Even
relatively modest contributions such as integrating support for specific models, such as DeepLabCut, or the more specific zebrafish feature-tracking package BonZeb, have attracted positive
feedback from the community, \citep[e.g.,][]{kaneEtAl20,guilbeaultEtAl21}.

### 2021-11-17 application/detailsOfResource.tex
merged

Bonsai is designed to be extensible and is distributed with a package-management framework that controls installation of optional core packages as well as user-contributed extensions.
%
However, extensions must be written in Bonsai's native C\# programming language.
%
This restricts the range of potential contributors to the ecosystem.
%
It also hinders incorporation of published analytic tools, many of which are implemented within other interpreted language systems such as Python, R, or MATLAB.  
\begin{wrapfigure}{R}{0pt}
  \includegraphics[width=0.5\textwidth]{figures/extensions.pdf}%
%   \input{figures/extensions}
  \caption{Proposed extensions to Bonsai}
  \label{fig:proposedBonsaiExtensions}
\end{wrapfigure}

We will extend the Bonsai package interface to allow direct interaction with the runtime execution environments that implement these common languages.  
%
Where possible, the extensions will be based on existing pipeline tools that allow C\#-based programs to interact with a number of alternative run-time engines.
%
These include:
%
Python.NET\footnote{https://github.com/pythonnet/pythonnet},
%
R.Net\break \footnote{https://github.com/rdotnet/rdotnet}
%
and the native .NET interface shipped with MATLAB.
% \footnote{https://uk.mathworks.com/help/matlab/matlab\_external/using-net-from-matlab-an-overview.html}.  TOO LONG
%
If necessary, we will also incorporate direct inter-runtime communication using the messaging library ZeroMQ\footnote{https://zeromq.org/}.

@@

    \item[Poisson Linear Dynamical System]\citep[PLDS][]{mackeEtAl15}: for
        smaller bin sizes, spike counts are better modelled as Poisson random
        processes, rather than Gaussian ones. The algorithm described in
        \citet{mackeEtAl15} can estimate the parameters of a LDS, and make
        inferences about its states, from Poisson distributed observations.  We
        propose to interface Bonsai with a Matlab implementation of
        PLDS\footnote{https://bitbucket.org/mackelab/ pop\_spike\_dyn/src/master/}
        that uses variational inference. PLDS does not provide online estimates
        of the states, since data from a full trial is needed for state
        inference.

@@

We propose assessing the progress of the proposed project using three 
types of targets.
%
First, we will have targets associated with each release to the general public
of the software components described above. 
% in Sections~\ref{sec:infrastructure} and~\ref{sec:functionality}.
%
Second, we will monitor user activity in the Bonsai
forum\footnote{https://groups.google.com/g/bonsai-users},
and publications citing Bonsai that use the new ML functionality,
to estimate the number of Bonsai users using this functionality.
%
Third, we will count the number of ML methods that are integrated
into the Bonsai ecosystem by us and by other method developers.

>>>>

Bonsai is designed to be extensible and is distributed with a package-management framework that controls installation of optional core packages as well as user-contributed extensions.
%
However, extensions must be written in Bonsai's native C\# programming language.
%
This restricts the range of potential contributors to the ecosystem.
%
It also hinders incorporation of published analytic tools, many of which are implemented within other interpreted language systems such as Python, R, or MATLAB.  
\begin{wrapfigure}{R}{0pt}
  \includegraphics[width=0.6\textwidth]{figures/extensions.pdf}%
%   \input{figures/extensions}
  \caption{Proposed extensions to Bonsai}
  \label{fig:proposedBonsaiExtensions}
\end{wrapfigure}

We will extend the Bonsai package interface to allow direct interaction with the runtime execution environments that implement these common languages.  
%
Where possible, the extensions will be based on existing pipeline tools that allow C\#-based programs to interact with a number of alternative run-time engines.
%
These include:
%
Python.NET\footnote{github.com/pythonnet/pythonnet},
%
R.Net\break \footnote{github.com/rdotnet/rdotnet}
%
and the native .NET interface shipped with MATLAB.
% \footnote{https://uk.mathworks.com/help/matlab/matlab\_external/using-net-from-matlab-an-overview.html}.  TOO LONG
%
If necessary, we will also incorporate direct inter-runtime communication using the messaging library ZeroMQ\footnote{zeromq.org/}.\\

@@

    \item[Poisson Linear Dynamical System]\citep[PLDS][]{mackeEtAl15}: for
        smaller bin sizes, spike counts are better modelled as Poisson random
        processes, rather than Gaussian ones. The algorithm described in
        \citet{mackeEtAl15} can estimate the parameters of a LDS, and make
        inferences about its states, from Poisson distributed observations.  We
        propose to interface Bonsai with a Matlab implementation of
        PLDS\footnote{bitbucket.org/mackelab/ pop\_spike\_dyn/ src/master/}
        that uses variational inference. PLDS does not provide online estimates
        of the states, since data from a full trial is needed for state
        inference.

@@

We propose assessing the progress of the proposed project using three 
types of targets.
%
First, we will have targets associated with each release to the general public
of the software components described above. 
% in Sections~\ref{sec:infrastructure} and~\ref{sec:functionality}.
%
Second, we will monitor user activity in the Bonsai
forum\footnote{groups.google.com/g/bonsai-users},
and publications citing Bonsai that use the new ML functionality,
to estimate the number of Bonsai users using this functionality.
%
Third, we will count the number of ML methods that are integrated
into the Bonsai ecosystem by us and by other method developers.

### 2021-11-17 application/narrativeCVRapela.tex
merged

Therefore, Dr.~Rapela is excellently trained to implement all aspects
(neuroscience, statistics, signal processing, and software development) of the
proposed project.

>>>>

Therefore, Dr.~Rapela's expertise is perfectly suited to implement all aspects
of the
proposed project (neuroscience, statistics, signal processing, and software development).

### 2021-11-17 application/otherResources.tex
merged

The large field of technologies serving experimental control and behaviour monitoring is traditionally occupied either by domain-specific graphical
user interfaces for control and recording of specific devices and experiment
types (e.g.\ Open Ephys
GUI\footnote{https://open-ephys.org/gui/},
Miniscope DAQ
Software\footnote{https://github.com/Aharoni-Lab/Miniscope-DAQ-QT-Software})
or by real-time control frameworks for specifying task logic using state
machine or similar formalisms (e.g. NIMH ML\footnote{monkeylogic.nimh.nih.gov/},
pyControl\footnote{pycontrol.readthedocs.io/ en/latest/},
Autopilot\footnote{docs.auto-pi-lot.com/en/latest/},
Sanworks\footnote{sanworks.io/index.php}).
These dedicated interfaces are typically very comfortable for experimenters in the specific domain for which the tool is designed, but can become unwieldy with the introduction of a new device or task variation from outside their usual scope. Alternatively, one can use a more general programming language such as Python or MATLAB, with the disadvantage
of the code being harder to understand, maintain, and change. Programming
languages like LabVIEW straddle the middle ground and provide a high-level, flexible visual interface for composing data acquisition and control systems. Unlike Bonsai, however, the graphical elements of LabVIEW are
heterogeneous and very fine grained, thus requiring long and complex
logical structures to implement even a simple experimental control system. By providing an extremely simple, yet flexible visual syntax, Bonsai provides the
opportunity for even complete non-programmers to design and successfully
customise relatively complex experiments from the ground up. It is this
capability in particular which has made Bonsai such an attractive standard tool in
experimental neuroscience.

>>>>

The large field of technologies serving experimental control and behaviour monitoring is traditionally occupied either by domain-specific graphical
user interfaces for control and recording of specific devices and experiment
types (e.g.\ Open Ephys
GUI\footnote{open-ephys.org/gui/},
Miniscope DAQ
Software\footnote{github.com/ Aharoni-Lab/Miniscope-DAQ-QT-Software})
or by real-time control frameworks for specifying task logic using state
machine or similar formalisms (e.g. NIMH ML\footnote{monkeylogic.nimh.nih.gov/},
pyControl\footnote{pycontrol. readthedocs.io/ en/latest/},
Autopilot\footnote{docs.auto-pi-lot.com/en/latest/},
Sanworks\footnote{sanworks.io/index.php}).\break
These dedicated interfaces are typically very comfortable for experimenters in the specific domain for which the tool is designed, but can become unwieldy with the introduction of a new device or task variation from outside their usual scope. Alternatively, one can use a more general programming language such as Python or MATLAB, with the disadvantage
of the code being harder to understand, maintain, and change. Programming
languages like LabVIEW straddle the middle ground and provide a high-level, flexible visual interface for composing data acquisition and control systems. Unlike Bonsai, however, the graphical elements of LabVIEW are
heterogeneous and very fine grained, thus requiring long and complex
logical structures to implement even a simple experimental control system. By providing an extremely simple, yet flexible visual syntax, Bonsai provides the
opportunity for even complete non-programmers to design and successfully
customise relatively complex experiments from the ground up. It is this
capability in particular which has made Bonsai such an attractive standard tool in
experimental neuroscience.
